Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-05 latex/main.tex
Batch of fixups

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ``X''), to which we refer as ``Pango X''.
Below, we summarize our analysis of the Pango Xs in the ARG
(for extended analyses, see Document S1 and Document S2).

We find strong evidence for the recombination events leading to 17 Pango Xs in
the ARG (Table~\ref{tab:pango_x_lineages}), all of which pass the 
robustness filtering criteria (Figure~\ref{fig:recombinant_qc}). 
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango Xs are XA, which
is the first designated Pango X that showed signs of community transmission
\cite{Jackson2021}, and XBB, which escaped vaccine-induced immunity and gained
global prevalence \cite{Tamura2023}. The recombinant origin of XA is well
explained in the ARG by two parents and one mutation (Figure \ref{fig:subgraphs};
Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and
breakpoint location.

Furthermore, some of the Pango Xs above originate from shared recombination events in the 
ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent events in the ARG (\ref{fig:subgraphs}).
These Pango Xs were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango Xs in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango Xs, the number of recombination-informative sites fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

@@

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
\label{tab:pango_x_lineages}}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
XAE+ & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
{BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {BA.2:4, XBH:2} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\ XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 4 & 1 \\
XAM & XQ & 6 & 43 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 55 & 4 & 6 \\
XAP & XAE+ & 4 & 67 & 1 & 20 \\
XZ & XAE+ & 4 & 48 & 1 & 48 \\
XAE & XAE+ & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 66 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}

% pango & averted & left & right & samples & nested \\

% \midrule
% XC & 37 & 26768 & 27390 & 5 & \{\} \\
% XBR & 22 & 22034 & 22190 & 1 & \{\} \\
% XA & 16 & 20411 & 21765 & 39 & \{\} \\
% XS & 11 & 9054 & 10449 & 17 & \{\} \\
% XL & 8 & 5925 & 8393 & 64 & \{\} \\
% XBG & 8 & 22305 & 22917 & 25 & \{\} \\
% XBD & 7 & 23020 & 24620 & 30 & \{\} \\
% XQ & 7 & 4322 & 5386 & 55 & \{BA.2:37, $\star$\} \\
% XBB & 6 & 22332 & 22577 & 6452 & \{BA.2:1\} \\
% XM & 6 & 17411 & 21595 & 26 & \{BA.2:16, XAL:3\} \\
% XBF & 5 & 5184 & 9866 & 185 & \{\} \\
% XF & 5 & 5387 & 6402 & 16 & \{\} \\
% XY & 5 & 11538 & 12880 & 23 & \{\} \\
% XG & 4 & 5925 & 6513 & 3 & \{\} \\
% XW & 4 & 2833 & 4321 & 32 & \{\} \\
% \midrule
% \multicolumn{5}{l}{Type II RE event}\\
% alias & averted & left & right & total & by pango \\
% \midrule
% XAE+ & 6 & 24504 & 26060 & 253 & {BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
% XE/XH & 6 & 10448 & 11283 & 1156 & {XE:1116, BA.2:37, XH:2, XAF:1} \\
% XBH & 6 & 15452 & 22001 & 6 & {BA.2:4, XBH:2} \\
% XBM & 6 & 22305 & 22917 & 12 & {XBM:10, BF.3:2} \\
% XAF & 5 & 10199 & 10447 & 36 & {BA.2:35, XAF:1} \\
% XJ & 5 & 13196 & 17410 & 85 & {XJ:68, BA.2:17} \\
% \midrule
% \multicolumn{5}{l}{Nested monophyletic Pango X lineage}\\
% pango & alias & path len & time & mutations & samples & nested \\
% \midrule
% XAP & XAE+ & 4 & 67 & 1 & 20 & \{\} \\
% XZ & XAE+ & 4 & 48 & 1 & 48 & \{\} \\
% XAE & XAE+ & 2 & 47 & 7 & 9 & \{\} \\
% XAF & XAF & 5 & 140 & 8 & 1 & \{\} \\
% XAL & XM & 3 & 66 & 2 & 3 & \{\} \\
% XR & XQ & 3 & 11 & 4 & 17 & \{BA.2:1\} \\
% XU & XQ & 6 & 103 & 4 & 1 & \{\} \\
% XAM & XQ & 6 & 43 & 3 & 21 & \{\} \\
% XAA & XQ & 6 & 33 & 3 & 17 & \{\} \\
% XAG & XQ & 6 & 55 & 4 & 6 & \{\} \\

% \bottomrule
% \end{tabular}

\end{table}

@@

To evaluate the concordance of the methods in identifying the parental Pango lineages
and locating the breakpoints of the recombinants, we focus on the Pango Xs, using the data
from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt, we took the detection results for the Pango Xs presented in the original study;
for CovRecomb, we summarized the results from the supplementary data of the original study (STAR Methods);
and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods).
For each method, we determined whether the inferred parental lineages and breakpoint intervals are
concordant with those identified by the community (STAR Methods).
We find that the methods performed well overall,
with no significant difference in concordance among them (Table 2; p-value = XX, chi-square test),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art recombination detection methods.

We also compared the sensitivity of sc2ts and rebar.
Of all the ``robust'' recombination events detected by sc2ts,
167 (43\%) were also detected as recombinants by rebar (STAR Methods).
We hypothesize that this disagreement arose from differences in sensitivity between these methods.
Consistent with this, we find that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by
a shorter time to their MRCA (\ref{fig:averted_mutations}) and
lower distance between their Pango identity (\ref{fig:averted_mutations}).

\subsection*{Table 2. Concordance among methods in detecting Pango Xs.}

\begin{table}
\begin{tabular}{llll}
 \hline
 \textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bp)} \\ [1ex]
 \hline
 Sc2ts & XX & XX & XX \\ [1ex]
 \hline
 RecombinHunt-GISAID & XX & XX & XX \\ [1ex]
 \hline
 RecombinHunt-Nextstrain & XX & XX & XX \\ [1ex]
 \hline
 CovRecomb & XX & XX & XX \\ [1ex]
 \hline
 Rebar & XX & XX & XX \\ [1ex]
 \hline
\end{tabular}
\end{table}

\subsection*{Quantification of evidence for recombination}

%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.

Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting mutations (\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination, the stronger the evidence for recombination.
We calculated the number of averted mutations for all the ``robust'' recombination events (STAR Methods),
which include all the Pango X-associated recombination events (Table 1).

Our analysis of the number of averted mutations reveals a spectrum of plausibility
across all the ``robust'' recombination events (\ref{fig:averted_mutations}).
Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1).
The ``copying pattern'' of XA visually confirms
that a high number of recombination-informative sites support the suggested parents (\ref{fig:averted_mutations}).
XW is an example of a marginally well-supported recombinant.
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
The copying patterns of all the recombination events are shown in Document S3.

Our results suggest that Delta and Omicron BA.2 might have recombinant origins,
but we interpret these results with caution,
as these VOCs are undersampled around the time of their emergence (REF).
Consistent with this, the presence of mutations and conflicting loci matching non-focal parents
observed in the copying patterns of these VOCs (Figure 3D) suggests the existence of
many genetically distinct, unsampled ancestors.
Future work is needed to test whether these VOCs descended from recombination events,
with improved sampling if still possible.

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
    \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
those identified by \cite{Jackson2021} (see Table \ref{tab:jackson}) are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
    \label{fig:averted_mutations}
\end{figure}

>>>>

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ``X''), to which we refer as ``Pango X''.
Below, we summarize our analysis of the Pango Xs in the ARG
(for extended analyses, see Document~\ref{sec:supplement_pango_lineages} 
and Document S2).

We find strong evidence for the recombination events leading to 17 Pango Xs in
the ARG (Table~\ref{tab:pango_x_lineages}), all of which pass the 
robustness filtering criteria (Figure~\ref{fig:recombinant_qc}). 
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango Xs are XA, which
is the first designated Pango X that showed signs of community transmission
\cite{Jackson2021}, and XBB, which escaped vaccine-induced immunity and gained
global prevalence \cite{Tamura2023}. The recombinant origin of XA is well
explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A;
Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and
breakpoint location.

Furthermore, some of the Pango Xs above originate from shared recombination events in the 
ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B).
These Pango Xs were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango Xs in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango Xs, the number of recombination-informative sites fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

@@

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
\label{tab:pango_x_lineages}}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
XAE+ & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
{BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {BA.2:4, XBH:2} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\ XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 4 & 1 \\
XAM & XQ & 6 & 43 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 55 & 4 & 6 \\
XAP & XAE+ & 4 & 67 & 1 & 20 \\
XZ & XAE+ & 4 & 48 & 1 & 48 \\
XAE & XAE+ & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 66 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

To evaluate the concordance of the methods in identifying the parental Pango lineages
and locating the breakpoints of the recombinants, we focus on the Pango Xs, using the data
from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt, we took the detection results for the Pango Xs presented in the original study;
for CovRecomb, we summarized the results from the supplementary data of the original study (STAR Methods);
and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods).
For each method, we determined whether the inferred parental lineages and breakpoint intervals are
concordant with those identified by the community (STAR Methods).
We find that the methods performed well overall,
with no significant difference in concordance among them
(Table~\ref{tab:x_method_concordance}; p-value = XX, chi-square test),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art recombination detection methods.

We also compared the sensitivity of sc2ts and rebar.
Of all the ``robust'' recombination events detected by sc2ts,
167 (43\%) were also detected as recombinants by rebar (STAR Methods).
We hypothesize that this disagreement arose from differences in sensitivity between these methods.
Consistent with this, we find that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by
a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
lower distance between their Pango identity (Figure
\ref{fig:averted_mutations}C).

\begin{table}
\caption{Concordance among methods in detecting Pango Xs.
\label{tab:x_method_concordance}}
\begin{tabular}{llll}
 \hline
 \textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bp)} \\ [1ex]
 \hline
 Sc2ts & XX & XX & XX \\ [1ex]
 \hline
 RecombinHunt-GISAID & XX & XX & XX \\ [1ex]
 \hline
 RecombinHunt-Nextstrain & XX & XX & XX \\ [1ex]
 \hline
 CovRecomb & XX & XX & XX \\ [1ex]
 \hline
 Rebar & XX & XX & XX \\ [1ex]
 \hline
\end{tabular}
\end{table}

\subsection*{Quantification of evidence for recombination}
%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting 
mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination, the stronger the 
evidence for recombination.
We calculated the number of averted mutations for all the ``robust'' recombination 
events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
al.\cite{Jackson2021}

Our analysis of the number of averted mutations reveals a spectrum of plausibility
across all the ``robust'' recombination events
(Figure~\ref{fig:averted_mutations}A).
Examples of highly plausible recombinants are XA and XBB (for an extended analysis, 
see Document~\ref{sec:supplement_pango_lineages}).
The ``copying pattern'' of XA visually confirms
that a high number of recombination-informative sites support the suggested parents 
(Figure~\ref{fig:averted_mutations}D).
XW is an example of a marginally well-supported recombinant.
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
% FIXME What's document S3?
The copying patterns of all the recombination events are shown in Document S3.

[FIXME - discuss BA.5 and the BA.2 event instead.]
Our results suggest that Delta and Omicron BA.2 might have recombinant origins,
but we interpret these results with caution,
as these VOCs are undersampled around the time of their emergence (REF).
Consistent with this, the presence of mutations and conflicting loci matching non-focal parents
observed in the copying patterns of these VOCs (Figure 3D) suggests the existence of
many genetically distinct, unsampled ancestors.
Future work is needed to test whether these VOCs descended from recombination events,
with improved sampling if still possible.

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
    \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
those identified by \cite{Jackson2021} (see Table \ref{tab:jackson}) are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
    \label{fig:averted_mutations}
\end{figure}
